# Spatio-temporal analysis of the effects of management strategies on the abundance of marine organisms

\noindent Diego R. Barneche

\doublespace \fontsize{12}{14}\selectfont
\setlength{\parindent}{15pt}

\noindent \textbf{Data accessibility statement:} All data and R code (data manipulation, analyses, figures and tables) can be downloaded from a GitHub repository (https://github.com/dbarneche/transectTask).

\noindent \textbf{Type of document:} Report

\noindent \textbf{\*Correspondence should be sent to:} Email: \href{mailto:barnechedr@gmail.com}{barnechedr@gmail.com}.

```{r, warnings = FALSE, eval = FALSE}
library(rmarkdown)
library(plyr)
library(bbmle)
library(glmmTMB)
library(brms)
library(effects)
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggsci)
library(png)
library(grid)
library(LoLinR) # devtools::install_github('colin-olito/LoLinR')
```

\clearpage

# The data

\noindent I am going to focus on a few analyses that are possible with the dataset in hand. First, we do a quick check to make sure that the data is clean.

```{r, warnings = FALSE, eval = FALSE}
source('R/analysis.R')
data  <-  readAndModifyData('data/data.csv')
```

```{r, warnings = FALSE}
str(data)
sapply(data, function (x) sum(is.na(x)))
```

\noindent So it looks like that apart from few NAs here and there, the dataset seems to be clean, i.e. without any typos nor mixed object classes within columns. 

# The main question

\noindent The main goal here is to investigate whether there is evidence for changes in fish abundance over time that could be attributable to two different management strategies which were put in place from 2004 onwards. This dataset comprises count data for four different species, and it has a strong nested spatial structure to its design. Transects will be considered as independent replicates within a given site, which in turn are nested in randomised blocks. Blocks are nested in regions. Transects are fixed on the substratum, and were revisited yearly from 2000 to 2014. So this is a typical longitudinal study.

There are two data explorations that are worth checking to begin with: (\textit{i}) whether there are any visible temporal trends in the data, and (\textit{ii}) how the data are distributed. Checking for any visually outstanding pattern below, the data shows there are no clear trends in either increase or decrease in abundance of different taxa (panel columns) across the four different regions (panel rows), and between the management strategies (circles of different colours). We could break it down into different blocks, and sites, but that would just create too many unnecessary figures---I will hopefully be able to capture the temporal dynamics and spatial variation with the chosen modelling exercise further down this text.

```{r, warnings = FALSE, eval = FALSE}
source('R/figures.R')
temporalTrends(data)
```
\clearpage 

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=1\textwidth]{../../output/figures/exploratoryTrends.png}
  \end{center}
  \caption*{}
\end{figure}


Count data are generally modelled with Poisson distribution, but can also be modelled with e.g. a Negative Binomial in case there is over-dispersion. Moreover, if zeros are exceedingly represented in the data, we can also model the data using Zero-Inflation or Hurdle approaches. With the former, we would be assuming that there are two processes producing zeros in the data, whereas with the latter we would be assuming that zeros are always generated by a single process that is distinct from the other process that generates the presences (i.e. non-zero counts). Here is a summary of data distributions broken down by species (columns in the data, and different panels in the figure), and region (different line types):

```{r, warnings = FALSE, eval = FALSE}
histograms(data)
```

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=1\textwidth]{../../output/figures/exploratoryHistograms.png}
  \end{center}
  \caption*{}
\end{figure}

We can see that there is strong zero-inflation in the data. Absences (i.e. zeros) in visual transects are more likely to occur because of both sampling bias, and a second process which could be inflating zeroes. Thus I will use a Zero-Inflated modelling approach in R [@rcitation] to take this possibility into account. I will do so using two packages: 1) `glmmTMB` [@glmmTMB] which implements Generalised Linear Mixed Models using Template Model Builder (TMB); and 2) `brms` [@brms] which fits models in a Bayesian framework using `stan` [@rstan]. I do so to make sure that the results converge between different approaches. However, given the available time, I establish the comparison explicitly for one species only---the Bayesian models tend to take many hours to run---and thereafter I will only report results using `glmmTMB`, Please visit the GitHub repository \href{https://github.com/dbarneche/transectTask/README.md}{README page} for a full specification of software and package versions used for this report.

# Methods

\noindent I chose to implement a Zero-Inflation model using a negative binomial distribution given the large over-dispersion in the data. Doing so ensured that convergence was properly achieved with both methods (`glmmTMB` and `brms`). This Zero-Inflated GLMM assumes a log-link function for the response, a log-link for the shape (over-dispersion) parameter, and a logit-link for the zero-inflated portion. Conveniently, the logit-link allows for the transformation of the `zi` statistic into a probability, which can be used as evidence (or lack thereof) for zero inflation.

The initial full model structure investigates changes in abundance over time, treating `Year` as a continuous covariate, with interactive effects attributable to `Management` (`a` or `b`), and a `baseline` effect on the intercept that could be attributable to pre- (i.e. before 2004) and post-implementation (2004--2014) of management strategies. `Site` were nested in `Block`, nested in `Region`, as random effects on the intercept. This approach allows the direct characterisation of variances for each one of these spatially nested levels, while maintaining the number of estimated parameters to a minimum---considering that each random effect variable contributes only one parameter to the model, i.e., its standard deviation. It is important to note, however, that regions only have 4 levels, and therefore this is really forcing the limit of hierarchical models, i.e. it is hard to reliably estimate a variance with four grouping levels.

I then either drop `baseline`, or the interaction between `Year` and `Managemet` (i.e. maintaining an additive effect only) from the full model, and compare the three models using AIC. Models with $\Delta$AIC < 2 were considered indistinguishable, and I kept the simplest one. It turns out that for the three species (damsel, wrasse, and parrotfish), all best models retained the interaction, but dropped the baseline effect (please recreate objects `bestTMBDamselModel`, `extractBestTMBModel`, `bestTMBWrasseModel` using the `remake` approach described on the \href{https://github.com/dbarneche/transectTask/README.md}{README page}---they were produced with the custom function `bestTMBParrotfishModel` in `R/analysis.R`). Model codes are as follows:

```{r, warnings = FALSE, eval = FALSE}
# damsel as an example
# in brms
brmsDamselModel  <-  brms::brm(damsel ~ baseline + Year * Management + 
  (1 | Region/Block/Site2), data = data, family = zero_inflated_negbinomial(),
  chains = 3)

# in glmmTMB
TMBDamselModel  <-  glmmTMB::glmmTMB(damsel ~ baseline + Year * Management +
  (1 | Region/Block/Site2), data = data, ziformula = ~1, family = nbinom2)
```

# Results

\noindent The species grouper was characterised by a different type of distribution not suitable for the Zero-Inflated modelling approach, and given the limited time availability I will report results for the remaining three species for which models have converged properly.

For the damsel species, in terms of approach, both packages yielded very similar results, indicating two main findings: 1) after accounting for potential residual autocorrelation attributable to the spatial nested design of the observations, abundances are decreasing with Management strategy `a`, and increasing with Management strategy `b`. However, 2) strategies are not substantially different from each other in terms of absolute values yet, as there is large uncertainty associated with the model estimated effects.

```{r, warnings = FALSE, eval = FALSE}
summary(brmsDamselModel)

##  Family: zero_inflated_negbinomial 
##   Links: mu = log; shape = identity; zi = identity 
## Formula: damsel ~ baseline + Year * Management + (1 | Region/Block/Site2) 
##    Data: data (Number of observations: 3569) 
## Samples: 3 chains, each with iter = 2000; warmup = 1000; thin = 1;
##          total post-warmup samples = 3000
## 
## Group-Level Effects: 
## ~Region (Number of levels: 4) 
##               Estimate Est.Error l-95% CI u-95% CI Eff.Sample Rhat
## sd(Intercept)     1.03      0.72     0.27     3.05        116 1.01
## 
## ~Region:Block (Number of levels: 20) 
##               Estimate Est.Error l-95% CI u-95% CI Eff.Sample Rhat
## sd(Intercept)     0.61      0.16     0.36     1.00         17 1.13
## 
## ~Region:Block:Site2 (Number of levels: 80) 
##               Estimate Est.Error l-95% CI u-95% CI Eff.Sample Rhat
## sd(Intercept)     0.54      0.07     0.43     0.68         66 1.03
## 
## Population-Level Effects: 
##                  Estimate Est.Error l-95% CI u-95% CI Eff.Sample Rhat
## Intercept           43.72      9.66    24.83    62.25        142 1.02
## baselineyes         -0.01      0.04    -0.09     0.06        107 1.04
## Year                -0.02      0.00    -0.03    -0.01        141 1.02
## Managementb        -87.25     10.55  -107.57   -65.65        134 1.01
## Year:Managementb     0.04      0.01     0.03     0.05        133 1.01
## 
## Family Specific Parameters: 
##       Estimate Est.Error l-95% CI u-95% CI Eff.Sample Rhat
## shape     5.66      0.25     5.20     6.17        136 1.02
## zi        0.44      0.01     0.42     0.46        129 1.02
## 
## Samples were drawn using sampling(NUTS). For each parameter, Eff.Sample 
## is a crude measure of effective sample size, and Rhat is the potential 
## scale reduction factor on split chains (at convergence, Rhat = 1).
## Warning message:
## The model has not converged (some Rhats are > 1.1). Do not analyse the results! 
## We recommend running more iterations and/or setting stronger priors. 

summary(TMBDamselModel)
##  Family: nbinom2  ( log )
## Formula: damsel ~ baseline + Year * Management + (1 | Region/Block/Site2)
## Zero inflation:          ~1
## Data: data
## 
##      AIC      BIC   logLik deviance df.resid 
##  18619.4  18681.2  -9299.7  18599.4     3559 
## 
## Random effects:
## 
## Conditional model:
##  Groups               Name        Variance Std.Dev.
##  Site2:(Block:Region) (Intercept) 0.2661   0.5158  
##  Block:Region         (Intercept) 0.3253   0.5704  
##  Region               (Intercept) 0.2136   0.4621  
## Number of obs: 3569, groups:  Site2:(Block:Region), 80;
## Block:Region, 20; Region, 4
## 
## Overdispersion parameter for nbinom2 family (): 5.68 
## 
## Conditional model:
##                    Estimate Std. Error z value Pr(>|z|)    
## (Intercept)       42.865470  10.250675   4.182 2.89e-05 ***
## baselineyes       -0.009952   0.040484  -0.246    0.806    
## Year              -0.020152   0.005105  -3.948 7.89e-05 ***
## Managementb      -86.475566  10.973365  -7.880 3.26e-15 ***
## Year:Managementb   0.043338   0.005471   7.921 2.35e-15 ***
## ---
## Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
## 
## Zero-inflation model:
##             Estimate Std. Error z value Pr(>|z|)    
## (Intercept) -0.23648    0.03567   -6.63 3.36e-11 ***
## ---
## Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
```

```{r, warnings = FALSE, eval = FALSE}
# from brms
managementPanelBrms(brmsDamselModel, picPath = 'pics/damsel.png', 
  ymin = log10(12), ymax = log10(60))
```

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=1\textwidth]{../../output/figures/damselManagementPanelBrms.png}
  \end{center}
  \caption*{}
\end{figure}


```{r, warnings = FALSE, eval = FALSE}
# from glmmTMB, also recreate bestTMBDamselModel using
#  `remake::make('bestTMBDamselModel')`
TMBModelsDamsel     <-  runTMBModels(data, response = 'damsel')
bestTMBDamselModel  <-  extractBestTMBModel(TMBModelsDamsel)
managementPanel(bestTMBDamselModel, data, picPath = 'pics/damsel.png', 
  ymin = 30, ymax = 38)
```

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=1\textwidth]{../../output/figures/damselManagementPanel.png}
  \end{center}
  \caption*{}
\end{figure}


Notice that some of the standard deviation estimates from the Bayesian approach did not converge properly, and this can be further illustrated by looking at the traceplots which show that some chains were not well mixed. However, given the caveat of lack of convergence, the model indicates a 44\% probability of that data being zero-inflated (`zi` statistic)---equivalent to $1 / (1 + e^{--0.23648})$ from `glmmTMB`---, and a considerable estimated over-dispersion (`shape` parameter = 5.68), strengthening the choice of this modelling approach using a negative binomial distribution. Variance between regions seems lower than within regions, which might indicate that the strongest spatial variation in the data occurs within regional levels.

The posterior distributions for the fixed effects seemed well sampled, and converged in mean estimates with the ones obtained using the package `glmmTMB`. These similarities although reassuring are perhaps not surprising because `brms` specifies weakly informative priors by default, causing the posterior distributions to converge to the likelihood estimates.

\clearpage 

```{r, warnings = FALSE, eval = FALSE}
posteriorDists(brmsDamselModel, pars = c('shape', 'zi', '^sd_'))
```

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=1\textwidth]{../../output/figures/damselPosteriorDistsRandom.png}
  \end{center}
  \caption*{}
\end{figure}


\clearpage 

```{r, warnings = FALSE, eval = FALSE}
posteriorDists(brmsDamselModel, pars = '^b_')
```

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=1\textwidth]{../../output/figures/damselPosteriorDistsFixed.png}
  \end{center}
  \caption*{}
\end{figure}


The `glmmTMB` analysis for wrasse and parrotfish also converged, however exhibited opposing management effects. While the wrasse showed similar trend to damsel (i.e. increasing in abundance under management strategy `b`), parrtofish seemed to be benefiting more from management strategy `a`.

```{r, warnings = FALSE, eval = FALSE}
TMBModelsWrasse     <-  runTMBModels(data, response = 'wrasse')
bestTMBWrasseModel  <-  extractBestTMBModel(TMBModelsWrasse)
TMBModelsParrotfish     <-  runTMBModels(data, response = 'parrotfish')
bestTMBParrotfishModel  <-  extractBestTMBModel(TMBModelsParrotfish)
managementPanel(bestTMBWrasseModel, data, picPath = 'pics/wrasse.png', 
  ymin = 28, ymax = 36)
managementPanel(bestTMBParrotfishModel, data, picPath = 'pics/parrotfish
  .png', ymin = 44, ymax = 49)
```

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.9\textwidth]{../../output/figures/wrasseManagementPanel.png}
  \end{center}
  \caption*{}
\end{figure}
\nointerlineskip
\begin{figure}[h]
  \begin{center}
    \includegraphics[width=0.9\textwidth]{../../output/figures/parrotfishManagementPanel.png}
  \end{center}
  \caption*{}
\end{figure}

# Main conclusions

\noindent \textit{Is there any evidence that management strategy effects the abundance of the focal species?} Yes, and differently so. Damsel and wrasse have increased in abundance over time under Management strategy `b`, whereas parrotfish has increased in abundance under Management strategy `a`.

\noindent \textit{At what spatial scale is variability in abundance and/or management effects greatest (are the management implications very local or do they have broader implications)?} Abundance within sites, as well as that between sites within blocks, varied more than when compared to regions. With this analysis, however, it is hard to say whether management is more effective locally or regionally. I have chosen to keep this analysis relatively simple and on the light-parametrised side, but further analysis with the addition of random slopes at the regional level could help elucidate this question.

\setlength{\parindent}{0pt}

# References
